---
title: "Module 4 Manuscript: Monica's Data"
subtitle: ""
author: Carter Coleman
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

The structure below is a possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed.


# Summary/Abstract
_Write a summary of your project._


# Introduction 

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_ 

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._



# Methods and Results

_In most research papers, results and methods are separate. You can combine them here if you find it easier. You are also welcome to structure things such that those are separate sections._


## Data aquisition
Data was obtained from the CDC's open source data archive on COVID-19 deaths during the COVID-19 pandemic.

## Data import and cleaning

#load needed packages. make sure they are installed.
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidyverse)
library(ggpubr)
library(AICcmodavg)
```

#path to data
#note the use of the here() package and not absolute paths
```{r}
data_location <- here::here("data","processed_data","processeddata.rds")
```

#load data. 
```{r}
NYC_Virus_Deaths <- readRDS(data_location)
```

#re-label missing data (NA) as 0. It is fair to assume no reported deaths as 0, as it is assumed death data is coming from hospitals, who must report all deaths.
```{r}
NYC_Virus_Deaths %>%
  replace_na(list(COVID.19.Deaths = 0, Influenza.Deaths = 0, Pneumonia.Deaths = 0))
```

## Exploratory analysis
#I'm using basic R commands here.
#Lots of good packages exist to do more.
#For instance check out the tableone or skimr packages

#summarize data 
```{r}
data_summary = summary(NYC_Virus_Deaths)
```

#look at summary
```{r}
print(data_summary)
```

#do the same, but with a bit of trickery to get things into the 
#shape of a data frame (for easier saving/showing in manuscript)
```{r}
summary_df = data.frame(do.call(cbind, lapply(NYC_Virus_Deaths, summary)))
```

#save data frame table to file for later use in manuscript
```{r}
summarytable_file = here("results", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

#Once the data is loaded, we will want to alter the month variable a bit to make it "month during pandemic". This will create a sequential variable data set, irrespective of year. This will help up later on.
```{r}
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 1] <- 13
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 2] <- 14
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 3] <- 15
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 4] <- 16
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 5] <- 17
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 6] <- 18
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 7] <- 19
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 8] <- 20
NYC_Virus_Deaths$Month[NYC_Virus_Deaths$Year == "2021"& NYC_Virus_Deaths$Month == 9] <- 21
```

#make a scatterplot of data (Covid-19 Deaths, Months)
#Make a separate data set for each of the sex-specific variables. 
```{r}
NVD_All <- NYC_Virus_Deaths %>% dplyr::filter(Sex == "All Sexes")
NVD_Female <- NYC_Virus_Deaths %>% dplyr::filter(Sex == "Female")
NVD_Male <- NYC_Virus_Deaths %>% dplyr::filter(Sex == "Male")
```

######################################
#Plotting Virus Deaths as a function of 
#Time, separated by gender and by total population (ALL)
######################################

#Plot newly formed data sets for covid deaths with line of best fit for each gender and combination.
```{r}
Covid_Month_All <- NVD_All %>% 
  ggplot(aes(x=Month, y=COVID.19.Deaths)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Covid Deaths per Month Since March 2020: all genders")
Covid_Month_Female <- NVD_Female %>% ggplot(aes(x=Month, y=COVID.19.Deaths)) + geom_point() + geom_smooth(method='lm')+ ggtitle("Covid Deaths per Month Since March 2020: Females")
Covid_Month_Male <- NVD_Male %>% ggplot(aes(x=Month, y=COVID.19.Deaths)) + geom_point() + geom_smooth(method='lm')+ ggtitle("Covid Deaths per Month Since March 2020: Males")
```

#look at each figure
```{r}
plot(Covid_Month_All)
plot(Covid_Month_Female)
plot(Covid_Month_Male)
```
##save figures
```{r}
#Total population
figure_file = here("results","COVID_ALL_figure.png")
ggsave(filename = figure_file, plot=Covid_Month_All)

#Females
figure_file = here("results","COVID_FEMALE_figure.png")
ggsave(filename = figure_file, plot=Covid_Month_Female)

#Males
figure_file = here("results","COVID_MALE_figure.png")
ggsave(filename = figure_file, plot=Covid_Month_Male)
```

##Repeat Previous steps for Influenza

#Plot newly formed data sets for Influenza deaths with line of best fit for each gender and combination.
```{r}
Influenza_Month_All <- NVD_All %>% ggplot(aes(x=Month, y=Influenza.Deaths)) + geom_point() + geom_smooth(method='lm') + ggtitle("Flu Deaths per Month Since March 2020: all genders")
Influenza_Month_Female <- NVD_Female %>% ggplot(aes(x=Month, y=Influenza.Deaths)) + geom_point() + geom_smooth(method='lm')+ ggtitle("Flu Deaths per Month Since March 2020: Females")
Influenza_Month_Male <- NVD_Male %>% ggplot(aes(x=Month, y=Influenza.Deaths)) + geom_point() + geom_smooth(method='lm')+ ggtitle("Flu Deaths per Month Since March 2020: Males")
```

#look at each figure
```{r}
plot(Influenza_Month_All)
plot(Influenza_Month_Female)
plot(Influenza_Month_Male)
```

##save figures
```{r}
#Total population

figure_file = here("results","Flu_ALL_figure.png")
ggsave(filename = figure_file, plot=Influenza_Month_All)

#Females
figure_file = here("results","Flu_FEMALE_figure.png")
ggsave(filename = figure_file, plot=Influenza_Month_Female)

#Males
figure_file = here("results","Flu_MALE_figure.png")
ggsave(filename = figure_file, plot=Influenza_Month_Male)
```

##Repeat Previous steps for Pneumonia

#Plot newly formed data sets for Influenza deaths with line of best fit for each gender and combination.
```{r}
Pneumonia_Month_All <- NVD_All %>% ggplot(aes(x=Month, y=Pneumonia.Deaths)) + geom_point() + geom_smooth(method='lm') + ggtitle("Pneumonia Deaths per Month Since March 2020: all genders")
Pneumonia_Month_Female <- NVD_Female %>% ggplot(aes(x=Month, y=Pneumonia.Deaths)) + geom_point() + geom_smooth(method='lm')+ ggtitle("Pneumonia Deaths per Month Since March 2020: Females")
Pneumonia_Month_Male <- NVD_Male %>% ggplot(aes(x=Month, y=Influenza.Deaths)) + geom_point() + geom_smooth(method='lm')+ ggtitle("Pneumonia Deaths per Month Since March 2020: Males")
```

#look at each figure
```{r}
plot(Pneumonia_Month_All)
plot(Pneumonia_Month_Female)
plot(Pneumonia_Month_Male)
```

##save figures

```{r}
#Total population
figure_file = here("results","Pneumonia_ALL_figure.png")
ggsave(filename = figure_file, plot=Pneumonia_Month_All)

#Females
figure_file = here("results","Pneumonia_FEMALE_figure.png")
ggsave(filename = figure_file, plot=Pneumonia_Month_Female)

#Males
figure_file = here("results","Pneumonia_MALE_figure.png")
ggsave(filename = figure_file, plot=Pneumonia_Month_Male)
```

##Finally, we will want to visualize how the three different virus deaths compare over time.
```{r}
NVD_ALL_2 <- pivot_longer(NVD_All, cols=6:8, names_to = "Virus_Type", values_to = "Deaths")

Virus_Death_Plot <- NVD_ALL_2 %>% 
  ggplot(aes(x=Month, y=Deaths, color = Virus_Type)) + 
  geom_point() + 
  geom_smooth(method = 'lm') + 
  ggtitle("Comparison of Virus Deaths Over Time")
```

#View combined virus death plot
```{r}
plot(Virus_Death_Plot)
```

#Save the figue
```{r}
figure_file = here("results","Virus_Deaths_figure.png")
ggsave(filename = figure_file, plot=Virus_Death_Plot)
```

## Full analysis

#Run an ANOVA test for the three different virus deaths over the entire provided time period.
```{r}
anova_one_way <- aov(Deaths~Virus_Type, data = NVD_ALL_2)
summary(anova_one_way)
```

# place results from fit into a data frame with the tidy function
```{r}
aov_table <- broom::tidy(anova_one_way)
```

 save fit results table  
```{r}
table_file_Covid = here("results", "Covid_resulttable.rds")
saveRDS(aov_table, file = table_file_Covid)
```

#look at fit results
```{r}
print(aov_table)
```

# Discussion and Conclusions
Based on the reported p-value of 0.382, there is no significant difference between deaths resulting from COVID, Influenza, or Pneumonia.

Based on previous knowledge about the high virulence of COVID-19 compared to Influenza and Pneumonia and the results of the produced figures, one possible explanation for this finding is reporting bias. COVID was the main focus of reported deaths during the pandemic. If you compare the created figures, n for Influenza and Pneumonia deaths is quite low. Low n suggests that the ANOVA performed was biased due to low sample size. Additionally, Pneumonia is a known secondary infection of both Influenza and Covid-19 and usually the cause of death when an individual dies from Covid. It is possible that there is a conflation of cause of death.


This paper [@Leek2015a] discusses types of analyses. 

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.


# References



